---
title: "Basic Deconvolution"
author: "Gregory Hunt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Deconvolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

We can show a basic deconvolution example by working with a data set created from rats. This is available on GEO with accession [GSE19830](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE19830).

First let's download the data set. 
```{r,warning=FALSE,cache=TRUE}
suppressPackageStartupMessages(library("affy"))
rat.url <- "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE19nnn/GSE19830/suppl/GSE19830_RAW.tar"
rat.dir <- "rat_data"
dir.create(rat.dir, showWarnings = FALSE)
rat.file <- "rat.tar"
download.file(rat.url, destfile = rat.file)
untar(rat.file, exdir = rat.dir)
rat <- ReadAffy(celfile.path = rat.dir)
```

This is a data set of rat brain, liver and lung cells that have been mixed together in various proportions and then analyzed with by DNA microarrays. The true mixing proportions are encoded in the following matrix. 

```{r}
rat.truth = matrix(c(0,0,0,1,1,1,0,0,0,0.25,0.25,0.25,0.05,0.05,0.05,0.7,0.7,0.7,0.25,0.25,0.25,0.45,0.45,0.45,0.2,0.2,0.2,0.3,0.3,0.3,0.3,0.3,0.3,0.4,0.4,0.4,0.35,0.35,0.35,0.34,0.34,0.34,1,1,1,0,0,0,0,0,0,0.05,0.05,0.05,0.7,0.7,0.7,0.25,0.25,0.25,0.7,0.7,0.7,0.45,0.45,0.45,0.55,0.55,0.55,0.5,0.5,0.5,0.55,0.55,0.55,0.5,0.5,0.5,0.6,0.6,0.6,0.65,0.65,0.65,0,0,0,0,0,0,1,1,1,0.7,0.7,0.7,0.25,0.25,0.25,0.05,0.05,0.05,0.05,0.05,0.05,0.1,0.1,0.1,0.25,0.25,0.25,0.2,0.2,0.2,0.15,0.15,0.15,0.1,0.1,0.1,0.05,0.05,0.05,0.01,0.01,0.01),ncol=3,byrow=FALSE)
head(rat.truth)
```

Each row is a sample and each column gives the mixing proportions of the cell type. From this we can extract out the pure samples of each of the three cell types. 
```{r}
pure_samples <- lapply(1:3, function(i) {
    which(rat.truth[, i] == 1)
})
pure_samples
```

Now the deconvolution algorithm can work with many different types of data. Ultimately we just need a matrix of expressions for each experiment. Let's first look at the probe level data. 
```{r,results='hide',echo=FALSE}
library('affy')
```
```{r}
Y <- log2(t(affy::intensity(rat)))
Y[1:4,1:4]
```
Here each row is a different experiment and each column is the log expression measurement of a particular oligonucleotide. 

Given the data and the pure samples we can run our deconvolution algorithm. 
```{r}
library('deconv')
n_choose <- 250
dc <- deconv(Y, pure_samples, n_choose)
phats <- dc$estimates
head(phats)
```
Here we have chosen to use the top 250 markers for each cell type. We plot the estimated mixing proportions against the truth.

```{r,results='asis',fig.height=5,fig.width=5}
plot(rat.truth,phats,xlab="Truth",ylab="Estimates")
abline(coef=c(0,1))
```


We can also run the algorithm on RMA summarized expressions. 
```{r,warning=FALSE,results='hide'}
Y <- data.frame(affy::rma(rat,verbose=FALSE))

n_choose <- 100
dc <- deconv(Y, pure_samples, n_choose)
phats <- dc$estimates
```
Here we have chosen to use the top 100 markers for each cell type. We plot the estimated mixing proportions against the truth.

```{r,results='asis',fig.height=5,fig.width=5}
plot(rat.truth,phats,xlab="Truth",ylab="Estimates")
abline(coef=c(0,1))
```
